---
title: "day10_ice"
author: "Steven Cognac"
date: "8/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(janitor)
library(ggbeeswarm)

library(GGally)

```


```{r}
ice <- read.csv(here("data", "ntl33_v7.csv"), na = -999)

temp <- read_csv(here("data", "ntl20_v6.csv"))
```
```{r}
ice_lakes <- ice %>% 
  group_by(lakeid) %>% 
  summarize(ice_duration)
  
ice_lakes

```

```{r}

ice_plot <- ggplot(ice_lakes, aes(x = lakeid, y = ice_duration, color = lakeid)) +
  geom_beeswarm() +
  geom_violin(alpha = 0.1)

ice_plot
```
Figure 1. Exploratory plot of ice duration for three lakes in the Madison Lake Area in Wisconsin. 

```{r}

## multiple observations per year because there are 3 sites.
ice_mean <- ice %>% 
  group_by(year4) %>% 
  summarize(ice_duration)

ice_mean

ice_mean_plot <- ggplot(ice_mean, aes(x = year4, y = ice_duration)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Year",
       y = "Ice duration",
       title = "Ice Duration in the Madison Lake Area over Time") +
  geom_smooth(method = "loess")

ice_mean_plot

```
Figure 2. Line plot of mean annual duration of ice cover in the Madison Lake Area of Wisconsin from 1851 to 2019. Long term trend shows a decline in the duration of ice cover over the period.

```{r}

mean_temp <- temp %>% 
  filter(month %in% c(1, 2, 12)) %>% 
  group_by(year4) %>% 
  summarize(mean_temp = mean(ave_air_temp_adjusted))
  



```

```{r}
winter_temp_plot <- ggplot(mean_temp, aes(x = year4, y = mean_temp)) +
  geom_line() +
  labs(x = "Year",
       y = "Temperature (C\U00B0)",
       title = "Average Winter Temperatures in the Madison, Wisconsin Lake Area",
       subtitle = "1851 - 2019") +
  theme_dark() +
  geom_smooth(method = "loess")

winter_temp_plot



```
Figure 3. Average December, January, and February temperatures in the Madison, Wisconsin Lake Area.  Long-term trends show a slight rise over time.

```{r}
ice_temp <- mean_temp %>% 
  full_join(ice_mean, by = "year4")



```

```{r}

ice_temp_plot <- ggplot(ice_temp, aes(x = ice_duration, y = mean_temp)) +
  geom_point(shape = 5, size = 0.5)

ice_temp_plot
```
Figure 4. Average ice duration compared to mean temperature (F).


```{r}
ice_thaw <- ice %>% 
  filter(year4 >= "1970") %>%
  mutate(Date = lubridate::ymd(ice_off)) %>% 
  mutate(day = lubridate::yday(Date))



```
```{r}
ice_off_plot <- ggplot(ice_thaw, aes(y = day, x = Date)) +
  geom_line() +
  labs(x = "Year",
       y = "Day of the Year",
       title = "Mean Ice Thaw Days in the Madison Lake Areas",
       subtitle = "1970 - 2019") +
  geom_smooth(method = "loess")

ice_off_plot

```
Figure 5. Average day of the year ice thaws in the Madison Lakes Area of Wisconsin.  Trend shows ice is thawing earlier in the year since the 1970s.